---
title: "Influence of UV irradiation on cSCC progression/metastasis assessed by epigenetic changes (miRNAs)"
output:
  html_notebook:
    theme: united
    toc: yes
author: Marc Bender
---

<style>
p.caption {
  font-size: 0.8em;
}
</style>

# Preamble

This document serves as a summary of the data analysis conducted during my PhD thesis. It is structured chronologically to emphasize decision making based on the previous results. The basic question was if UV radiation influences tumor growth beyond initiation e.g. by induction of genomic changes or epigenetic alterations resulting in tumor progression and metastasis of cutaneous squamous cell carcinoma (cSCC). Data is available at: https://github.com/MBender1992/PhD/blob/Marc/Data/200619_chronic_irr_normalized.csv

### Loading of required packages

```{r}
# data wrangling and basic visualization
library(tidyverse)
library(ggpubr)
# pipe-friendly statistics
library(rstatix)
# Heatmap
library(ComplexHeatmap)
library(circlize)
# PCA
library(FactoMineR)
library(factoextra)
```


# 1. Screening for UV induced miRNAs
## 1.1 Experimental setup and data ingestion

In a first step we screened 51 manually curated miRNAs associated with cSCC or other epithelial tumors for UV induced changes. To this end, five different cancer cell lines (SCL-II, Met-1, Met-4, SCC-12, SCC-13; Fig. 1) were grown under standard cell culture conditions with DMEM (Met-1,Met-4, SCL-II) or FAD (SCC-12, SCC-13) as cell culture media. Cells were grown to 80 % confluency and exposed to 103 kJ/m² (physical doses: 246.01 J/m² UVB (280-315 nm); 12.50 kJ/m² UVA (315-400 nm); 28.03 kJ/m² VIS (400-750 nm); 61.74 kJ/m² IR-A (750-1200 nm)) sunlight-equivalent radiation (SER) with a custom-built irradiation device (KAUVIR: https://www.fona.de/de/massnahmen/foerdermassnahmen/strahlenforschung.php). For the analysis 4 biological replicates were used.  


![**Fig. 1: .**  )]()


After irradiation, cells were harvested and subjected to FirePlex immunoassay (abcam;  https://www.abcam.com/nav/multiplex-assays/fireplex-immunoassays) to detect changes in miRNA expression using barcode-labelled probes. Data was imported to the complimentary workbench software and normalized to the 12 most stable miRNAs using the built-in algorithm.


```{r}
normalizers <- c("miR-23a-3p","miR-23b-3p", "miR-24-3p", "miR-26a-3p", "miR-30b-3p", "miR-22-3p", "miR-16-5p", "miR-221-3p", "miR-15a-3p", "miR-29a-3p", "let-7f-5p", "miR-29c-3p")
normalizers
```
   
Subsequently, data was imported into R using the following function, to exclude miRNAs with a median expression below an arbitrary threshold, convert data from wide to tidy data and add a column with log2-transformed values as expression data follows a log-normal distribution (Beal 2017, https://ieeexplore.ieee.org/stamp/stamp.jsp?arnumber=7991639) :   

```{r, message = FALSE}
load_Fireplex_data_PhD <- function(filename, threshold)
{
  # check if required package is installed and load it
  if (!require(dplyr)) install.packages("dplyr")
  library(dplyr)
  
  #load data and change hyphens to underscores so r does not confuse a hyphen with minus
  dat <- read_csv(filename)
  names(dat) <- str_replace_all(names(dat), "-","_")
  # clean up the data, convert to tidy format and replace negative expression values with 0 as expression        cannot be negative
  dat_f <- dat %>% 
    select(-Messung) %>%
    filter(cell_line != "HaCaT") %>%
    gather(miRNA, expression, -c(ID,cell_line, Irradiation)) %>% 
    mutate(expression = ifelse(expression < 0, 0, expression))  
  # index miRNAs with a median expression below the threshold
  ind <- dat_f %>%
    group_by(miRNA) %>%
    filter(median(expression, na.rm=TRUE) <= threshold) %>%
    .$miRNA %>% 
    unique() 
  # omit indexed miRNAs and add log2-transformed expression values 
  dat_f %>% 
    filter(!miRNA %in% ind) %>%
    mutate(log_exp = log2(expression+1)) %>%
    mutate(miRNA = str_replace_all(miRNA, "hsa_", "")) %>%
    mutate(miRNA = str_replace_all(miRNA, "_","-")) %>%
    mutate(cell_line = str_replace_all(cell_line, "-","_"))
}

url_file <- "https://raw.githubusercontent.com/MBender1992/PhD/Marc/Data/200619_chronic_irr_normalized.csv" 
dat <-  load_Fireplex_data_PhD(filename = url(url_file), threshold = 2.5)
dat
```

The threshold was set to 2.5 (a.u.) as this is the detection limit of the assay as indicated by the manufacturers instructions. After applying this threshold 40 out of the intially 51 examined miRNAs remained in the analysis. 

## 1.2 Principal component analysis (PCA)

Afterwards, data was analyzed for general changes between cell lines. As explorative analysis,PCA was chosen as it breaks down multidimensional data and facilitates visualization of the underlying data structure. 

```{r}
# transform data for the format required by "PCA"
dat_PCA <- dat %>% 
  select(-c(expression)) %>%
  spread(miRNA, log_exp) %>%
  mutate(cell_line = str_replace_all(cell_line, "_","-"))

# scale data and perform PCA
res.pca <- PCA(dat_PCA[, -c(1:3)],  scale.unit = T, ncp = 5, graph = F)

# extract results and chose which PCs to use
eig.val <- get_eigenvalue(res.pca)

head(eig.val)

```

As shown above, when using three dimensions, the analysis is able to explain more than 70 % of the variance in the data. As we add more dimensions the explanation improves, naturally, however for visualization reasons and the complexity of imagining more than three dimensions, only the comparison of PC1 vs PC2, PC1 vs PC3 and PC2 vs PC3 are shown (Fig. 1).  

```{r}
fviz_pca_ind(
  res.pca,
  geom.ind = "point",
  fill.ind = dat_PCA$cell_line,
  col.ind = "black",
  pointshape = 21, 
  pointsize = dat_PCA$Irradiation,
  palette = "jco",
  addEllipses = T,
  axes = c(1, 2),
  # Variables
  alpha.var =0.5,
  mean.point = F,
  ellipse.level=0.8,
  legend.title = list(fill = "Cell line")
) + 
  scale_size_manual(values = c(2,4))
```
![**Fig. 2: PCA of the irradiated cell lines Met-1, Met-4, SCC-12, SCC-13, SCL-II.** Large circles show irradiated samples, small circles indicate control samples. Dim1 is equivalent to Principal Component 1 (PC1), etc. Ellipses are drawn around the respective cell lines for illustration. The variance explained is shown in parentheses. NOTE: The code above only shows the plot for PC1 vs PC2 (axes = c(1,2)). The image was generated by saving the 3 different images (A: PC1 vs PC2,B: PC1 vs PC3,C: PC2 vs PC3) and combining them in power point. ](https://raw.githubusercontent.com/MBender1992/PhD/Marc/Results/PCA_combined.png)


A separation of SCC-12/SCC-13 and Met-1/Met-4 is evident along the PC1 axis, PC2 separates SCL-II from all other cell lines. Within clusters (ellipses) irradiated cells (big circles) form subclusters in the PC1 vs PC2 plot as well as in the PC2 vs PC3 plot in all cell lines and in the PC1 vs PC3 plot in Met-4, SCC-13, and SCL-II. However the irradiation effects are not as pronounced as the cell line effects indicating a larger difference in the miRNA expression between cell lines than between treated and control cells. 

To further investigate the overall irradiation effect a second PCA plot has been drawn to emphasize differences in irradiated and control cells (Fig. 2).

```{r}
fviz_pca_biplot(
  res.pca,
  # Fill individuals by groups
  geom.ind = "point",
  pointshape = 21,
  pointsize = 2.5,
  fill.ind = dat_PCA$Irradiation,
  col.ind = "black",
  palette = "jco",
  # Color variable by groups
  col.var = "cos2",
  alpha.var = 0.3,
  axes = c(1,2),
  addEllipses = T,
  ellipse.level=0.8,
  legend.title = list(fill = "Irradiation", color = "quality of representation"),
  repel = TRUE        # Avoid label overplotting
)+ 
  scale_size_manual(values = c(2,4))
```
![**Fig. 3: PCA of the irradiated cell lines Met-1, Met-4, SCC-12, SCC-13, SCL-II.** Large circles represent the mean. Dim1 is equivalent to Principal Component 1 (PC1), etc. Ellipses are drawn around the treatment conditions. Arrows indicate correlation of miRNA with the respective Principal Component. miRNAs sharing a similar angle in the plot are positively correlated, miRNAs with orthogonal arrows are uncorrelated and miRNAs with an angle of 180° are negatively correlated. The variance explained is shown in parentheses. NOTE: The code above only shows the plot for PC1 vs PC2 (axes = c(1,2)). The image was generated by saving the 3 different images (A: PC1 vs PC2,B: PC1 vs PC3,C: PC2 vs PC3) and combining them in power point. ](https://raw.githubusercontent.com/MBender1992/PhD/Marc/Results/PCA_combined_irradiation.png)

Fig. 2 shows the irradiation effect correlating with PC2 and PC3.Taken together these finding suggest the majority of variance in the data originating in cell line differences as indicated by the separation of cell lines by PC1 and a smaller but still substantial portion of the variance stemming from the irradiation treatment. These two effects - *cell line effect* and *irradiation effect* - were further analyzed. 

## 1.3 Hierarchical clustering analysis of the cell line effect

To get a better understanding of the cell line effect a hierarchical clustering analysis on differentially expressed miRNAs in control cells was performed.The statistical analysis underlying the characterization as *differentially expressed* is shown as follows: 


### 1.3.1 Checking assumptions for statistical tests
  

```{r}
# check for outliers 
outl <- dat %>% 
  group_by(cell_line,Irradiation, miRNA) %>%
  identify_outliers(log_exp) 
any(outl$is.extreme)
```

No extreme outliers were present in the analysis. Testing for normality or homogeneity of variances was not indicated in this study as the sample size was too small (n = 4) to detect deviations from said assumptions. Therefore,  a theoretical approach has been used to solve this problem:

* As explained in section 1.1, expression data follows a log-normal distribution, thus log-transformed values were used for statistical analysis 
* To deal with possible heterogeneity of variances, ANOVA with heteroscedasticity corrected covariance matrices (White 1980, Long and Ervin 2000) and post-hoc pairwise t-test with unequal variances were employed (white.adjust = TRUE in ANOVA and pool.sd = TRUE in pairwise t-test)

### 1.3.2 Two-Way ANOVA

A two-way ANOVA was conducted at each level of the factor *miRNA* to determine the effects of cell line and irradiation on miRNA expression. Statistical significance was accepted for p < 0.05 after fdr-adjustment (Benjamini & Hochberg 1995) for two-way interactions and simple main effects. The results of the two-way ANOVA are shown exemplary below.

STR-SHIFT + Enter --> code chunk + output
```{r, message=FALSE}
# 2-way ANOVA with miRNA as moderator variable
two_way_ANOVA <- dat %>%
  group_by(miRNA) %>%
  anova_test(log_exp ~ cell_line*Irradiation, white.adjust = T) %>%
  adjust_pvalue(method = "fdr") 
two_way_ANOVA
```

In 5 out of the 40 miRNAs there was an interaction effect present (miR-126-3p, miR-146a-5p, miR-30a-3p, miR-30b-5p, miR-7-5p). A cell line main effect was observed in 38 out of 40 miRNAs (each of the miRNAs showing an interaction effect also showed a sigfnifcant main effect). Only for miRNA let-7f-p and miR-30d-5p no cell line effect could be observed after adjusting for multiple testing. 


```{r}
# show interaction
two_way_interaction <- two_way_ANOVA %>%
  filter(Effect == "cell_line:Irradiation" & p.adj < 0.05) %>%
  pull(miRNA)
two_way_ANOVA %>% 
  filter(miRNA %in% two_way_interaction & Effect == "cell_line:Irradiation") 

# show main effect
two_way_main<- two_way_ANOVA %>%
  filter(Effect == "cell_line" & p.adj < 0.05) %>%
  pull(miRNA)
two_way_ANOVA %>% 
  filter(miRNA %in% two_way_main & Effect == "cell_line") 

# no effect
two_way_ANOVA %>% 
  filter(!miRNA %in% two_way_main & Effect == "cell_line")
```

### 1.3.3 One-Way ANOVA

To further investigate the cell line effect a one-way ANOVA was conducted with miRNA and Irradiation as grouping variables. To eliminate the Irradiation effect and only compare *baseline* miRNA expression between cell lines the results were filtered for control cells only, resulting in 36 differentially expressed miRNAs.


```{r, message=FALSE}
# 1-way ANOVA 
one_way_ANOVA <- dat %>%
  group_by(miRNA, Irradiation) %>%
  anova_test(log_exp ~ cell_line, white.adjust = T) %>%
  adjust_pvalue(method = "fdr") %>% 
  filter(Irradiation == "control" & p.adj < 0.05) %>%
  select(-Irradiation)
one_way_ANOVA
```

*NOTE: using the pooled variance from a 2-way ANOVA in the following 1-way ANOVA yields a higher statistical power but is more susceptible to violations from the homogeneity of variance assumptions. Analogously post-hoc tests with pooled variances (pool.sd = TRUE) are better in situations when the assumptions are met, at the cost of being less robust when assumptions not met.*

### 1.3.4 Heatmap

The differentially expressed miRNAs were used in a hierarchical clustering analysis and visualized in a Heatmap to reduce complexity and show the quantitative properties of each miRNA in an overview approach (Fig. 3). 


```{r}
# Heatmap  
# transform data for Heatmap format
dat_Heatmap <- dat %>% 
  filter(Irradiation == "control" & miRNA %in% pwc_cell_line$miRNA) %>%
  select(-c(Irradiation, log_exp)) %>%
  spread(miRNA, expression) %>%
  mutate(cell_line = str_replace_all(cell_line, "_","-"))

# scale data for Heatmap
dat_scaled <- dat_Heatmap %>% 
  column_to_rownames("ID") %>% 
  select(-cell_line) %>%
  scale() %>%
  t()

# set ID as rownames so colorbar works properly
dat_colorbar <- dat_Heatmap %>% select(c(ID, cell_line)) %>% 
  column_to_rownames("ID")

# define colors for colorbar
colorbar <- HeatmapAnnotation(
  df =dat_colorbar,
  col = list(
    cell_line = c(
      "Met-1" = "#0073C2FF",
      "Met-4" = "#EFC000FF",
      "SCC-12"="#868686FF" ,
      "SCC-13"="#CD534CFF",
      "SCL-II"="#7AA6DCFF")
    ),
  annotation_legend_param = list(
    cell_line = list(nrow=1)
    )
  )

# define color for the Heatmap cells
col_fun = colorRamp2(c(-1.8, 0, 1.8), c(c("#6D9EC1", "white", "#E46726")))

# Draw Heatmap 
Ht <- Heatmap(
  dat_scaled,
  col= col_fun,
  top_annotation = colorbar,
  row_split = 4,
  column_title = c("A", "B", "C"),
  border = T,
  column_km = 3, #  split cell lines in 3 clusters according to PCA
  column_km_repeats = 100, # repeat km split 100 times to get a stable cluster
  clustering_method_row = "average", # average linkage
  clustering_method_columns = "average",
  clustering_distance_row = "pearson", 
  clustering_distance_column = "euclidean",
  rect_gp = gpar(col = "white",lty = 1, lwd = 1),
  row_names_gp = gpar(fontsize = 10),
  show_column_names = FALSE,
  column_names_gp = gpar(fontsize = 10),
  heatmap_legend_param = list(
    title = "row Z-score",
    at = seq(-2,2,by=1),
    color_bar="continuous",
    title_position ="topcenter",
    legend_direction = "horizontal",
    legend_width = unit(4, "cm")
    ))

draw(Ht, annotation_legend_side = "bottom", heatmap_legend_side = "bottom")
```

![**Fig. 4: Hierarchical clustering analysis of differentially expressed miRNAs between the five different cell lines.** K-means clustering was applied with k=3 splits as PCA indicated 3 distinct clusters (Met-1/Met-4, SCC-12/SCC-13 and SCL-II). These clusters are also present in the Heatmap displayed by the color code below the column dendrogram. Orange shows upregulation of miRNA expression, blue shows downregulation of miRNA expression. Clustering method: average linkage. Distance measure for miRNAs: pearson correlation. Distance measure for cell lines: euclidean distance.  )](https://raw.githubusercontent.com/MBender1992/PhD/Marc/Results/Heatmap_cell_lines.png)

Supervised k-means cluster analysis showed a clear separation of miRNA expression depending on the cell line context (A: SCC-12/SCC-13,B: SCL-II,C: Met-1/Met-4).
The miRNAs were clustered in 4 groups. Cluster 1 was upregulated in cell line cluster A compared to cluster B and C. Cluster 2 was upregulated in cell line cluster B compared to cluster A and C. Cluster 4 was upregulated in cell line cluster C compared to A and B. Cluster 3 showed an intermediate expression of miRNAs in Cluster B and C with the lowest expression values observed in Cluster A.

Taken together these findings indicate distinct miRNA clusters representing a cell line specific epigenetic pattern 

