---
title: "Influence of UV irradiation on cSCC progression/metastasis assessed by epigenetic changes (miRNAs)"
output:
  html_notebook:
    theme: united
    toc: yes
author: Marc Bender
---

<style>
p.caption {
  font-size: 0.8em;
}
</style>

# Preamble

This document serves as a summary of the data analysis conducted during my PhD thesis. It is structured chronologically to emphasize decision making based on the previous results. The basic question was if UV radiation influences tumor growth beyond initiation e.g. by induction of genomic changes or epigenetic alterations resulting in tumor progression and metastasis of cutaneous squamous cell carcinoma (cSCC). Data is available at: https://github.com/MBender1992/PhD/blob/Marc/Data/200619_chronic_irr_normalized.csv

### Loading of required packages

```{r, message = FASLE}
# data wrangling and basic visualization
library(tidyverse)
library(ggpubr)
library(data.table)
# statistics
library(rstatix)
library(EnvStats) 
# Heatmap
library(ComplexHeatmap)
library(circlize)
# PCA
library(FactoMineR)
library(factoextra)
```


# 1. Screening for UV induced miRNAs
## 1.1 Experimental setup and data ingestion

In a first step we screened 51 manually curated miRNAs associated with cSCC or other epithelial tumors for UV induced changes. To this end, five different cancer cell lines (SCL-II, Met-1, Met-4, SCC-12, SCC-13; Fig. 1) were grown under standard cell culture conditions with DMEM (Met-1,Met-4, SCL-II) or FAD (SCC-12, SCC-13) as cell culture media. Cells were grown to 80 % confluency and exposed to 103 kJ/m² (physical doses: 246.01 J/m² UVB (280-315 nm); 12.50 kJ/m² UVA (315-400 nm); 28.03 kJ/m² VIS (400-750 nm); 61.74 kJ/m² IR-A (750-1200 nm)) sunlight-equivalent radiation (SER) with a custom-built irradiation device (KAUVIR: https://www.fona.de/de/massnahmen/foerdermassnahmen/strahlenforschung.php). The cells were treated eight times with this dose over time course of 4 weeks to simulate a chronic UV exposure. For the analysis 4 biological replicates were used.  


![**Fig. 1: Cell line characteristics of the tumor cell lines used in this study.**  The cell lines exhibit different levels of differentiation with SCC-13 being highly differentiated and SCL-II being undifferentiated. p53 mutations are present in all cell lines except Met-1 and Met-4. Met-1 and Met-4 are cell lines from the same patient in different stages of the tumor.)](https://raw.githubusercontent.com/MBender1992/PhD/Marc/Results/cell_line_characterization_cropped.png)


After irradiation, cells were harvested and subjected to FirePlex immunoassay (abcam;  https://www.abcam.com/nav/multiplex-assays/fireplex-immunoassays) to detect changes in miRNA expression using barcode-labelled probes. Data was imported to the complimentary workbench software and normalized to the 12 most stable miRNAs using the built-in algorithm.


```{r}
normalizers <- c("miR-23a-3p","miR-23b-3p", "miR-24-3p", "miR-26a-3p", "miR-30b-3p", "miR-22-3p", "miR-16-5p", "miR-221-3p", "miR-15a-3p", "miR-29a-3p", "let-7f-5p", "miR-29c-3p")
normalizers
```
   
Subsequently, data was imported into R using the following function, to exclude miRNAs with a median expression below an arbitrary threshold, convert data from wide to tidy data and add a column with log2-transformed values as expression data follows a log-normal distribution (Beal 2017, https://ieeexplore.ieee.org/stamp/stamp.jsp?arnumber=7991639) :   

```{r, message = FALSE}
load_Fireplex_data_PhD <- function(filename, threshold)
{
  # check if required package is installed and load it
  if (!require(dplyr)) install.packages("dplyr")
  library(dplyr)
  
  #load data and change hyphens to underscores so r does not confuse a hyphen with minus
  dat <- read_csv(filename)
  names(dat) <- str_replace_all(names(dat), "-","_")
  # clean up the data, convert to tidy format and replace negative expression values with 0 as expression        cannot be negative
  dat_f <- dat %>% 
    select(-Messung) %>%
    filter(cell_line != "HaCaT") %>%
    gather(miRNA, expression, -c(ID,cell_line, Irradiation)) %>% 
    mutate(expression = ifelse(expression < 0, 0, expression))  
  # index miRNAs with a median expression below the threshold
  ind <- dat_f %>%
    group_by(miRNA) %>%
    filter(median(expression, na.rm=TRUE) <= threshold) %>%
    .$miRNA %>% 
    unique() 
  # omit indexed miRNAs and add log2-transformed expression values 
  dat_f %>% 
    filter(!miRNA %in% ind) %>%
    mutate(log_exp = log2(expression+1)) %>%
    mutate(miRNA = str_replace_all(miRNA, "hsa_", "")) %>%
    mutate(miRNA = str_replace_all(miRNA, "_","-")) %>%
    mutate(cell_line = str_replace_all(cell_line, "-","_"))
}

url_file <- "https://raw.githubusercontent.com/MBender1992/PhD/Marc/Data/200619_chronic_irr_normalized.csv" 
dat <-  load_Fireplex_data_PhD(filename = url(url_file), threshold = 2.5)
dat
```

The threshold was set to 2.5 (a.u.) as this is the detection limit of the assay as indicated by the manufacturers instructions. After applying this threshold 40 out of the intially 51 examined miRNAs remained in the analysis. 

## 1.2 Principal component analysis (PCA)

Afterwards, data was analyzed for general changes between cell lines. As explorative analysis,PCA was chosen as it breaks down multidimensional data and facilitates visualization of the underlying data structure. 

```{r}
# transform data for the format required by "PCA"
dat_PCA <- dat %>% 
  select(-c(expression)) %>%
  spread(miRNA, log_exp) %>%
  mutate(cell_line = str_replace_all(cell_line, "_","-"))

# scale data and perform PCA
res.pca <- PCA(dat_PCA[, -c(1:3)],  scale.unit = T, ncp = 5, graph = F)

# extract results and chose which PCs to use
eig.val <- get_eigenvalue(res.pca)

head(eig.val)

```

As shown above, when using three dimensions, the analysis is able to explain more than 70 % of the variance in the data. As we add more dimensions the explanation improves, naturally, however for visualization reasons and the complexity of imagining more than three dimensions, only the comparison of PC1 vs PC2, PC1 vs PC3 and PC2 vs PC3 are shown (Fig. 2).  

```{r}
fviz_pca_ind(
  res.pca,
  # Fill individuals by groups
  geom.ind = "point",
  fill.ind = dat_PCA$cell_line,
  col.ind = "black",
  pointshape = 21, 
  pointsize = dat_PCA$Irradiation,
  palette = "jco",
  addEllipses = T,
  axes = c(1, 2),
  # Variables
  alpha.var =0.5,
  mean.point = F,
  ellipse.level=0.8,
  legend.title = list(fill = "Cell line")
) + 
  scale_size_manual(values = c(2,4))
```
![**Fig. 2: PCA of the irradiated cell lines Met-1, Met-4, SCC-12, SCC-13, SCL-II.** Large circles show irradiated samples, small circles indicate control samples. Dim1 is equivalent to Principal Component 1 (PC1), etc. Ellipses are drawn around the respective cell lines for illustration. The variance explained is shown in parentheses. NOTE: The code above only shows the plot for PC1 vs PC2 (axes = c(1,2)). The image was generated by saving the 3 different images (A: PC1 vs PC2,B: PC1 vs PC3,C: PC2 vs PC3) and combining them in power point. n = 4.](https://raw.githubusercontent.com/MBender1992/PhD/Marc/Results/PCA_combined.png)


A separation of SCC-12/SCC-13 and Met-1/Met-4 is evident along the PC1 axis, PC2 separates SCL-II from all other cell lines. Within clusters (ellipses) irradiated cells (big circles) form subclusters in the PC1 vs PC2 plot as well as in the PC2 vs PC3 plot in all cell lines and in the PC1 vs PC3 plot in Met-4, SCC-13, and SCL-II. However the irradiation effects are not as pronounced as the cell line effects indicating a larger difference in the miRNA expression between cell lines than between treated and control cells. 

To further investigate the overall irradiation effect a second PCA plot has been drawn to emphasize differences in irradiated and control cells (Fig. 3).

```{r}
fviz_pca_biplot(
  res.pca,
  # Fill individuals by groups
  geom.ind = "point",
  pointshape = 21,
  pointsize = 2.5,
  fill.ind = dat_PCA$Irradiation,
  col.ind = "black",
  palette = "jco",
  # Color variable by groups
  col.var = "cos2",
  alpha.var = 0.3,
  axes = c(1,2),
  addEllipses = T,
  ellipse.level=0.8,
  legend.title = list(fill = "Irradiation", color = "quality of representation"),
  repel = TRUE        # Avoid label overplotting
)+ 
  scale_size_manual(values = c(2,4))
```
![**Fig. 3: PCA of the irradiated cell lines Met-1, Met-4, SCC-12, SCC-13, SCL-II.** Large circles represent the mean. Dim1 is equivalent to Principal Component 1 (PC1), etc. Ellipses are drawn around the treatment conditions. Arrows indicate correlation of miRNA with the respective Principal Component. miRNAs sharing a similar angle in the plot are positively correlated, miRNAs with orthogonal arrows are uncorrelated and miRNAs with an angle of 180° are negatively correlated. The variance explained is shown in parentheses. NOTE: The code above only shows the plot for PC1 vs PC2 (axes = c(1,2)). The image was generated by saving the 3 different images (A: PC1 vs PC2,B: PC1 vs PC3,C: PC2 vs PC3) and combining them in power point. n = 4. ](https://raw.githubusercontent.com/MBender1992/PhD/Marc/Results/PCA_combined_irradiation.png)

Fig. 3 shows the irradiation effect correlating with PC2 and PC3.Taken together these finding suggest the majority of variance in the data originating in cell line differences as indicated by the separation of cell lines by PC1 and a smaller but still substantial portion of the variance stemming from the irradiation treatment. These two effects - *cell line effect* and *irradiation effect* - were further analyzed. 

## 1.3 Investigation of the cell line effect

To get a better understanding of the cell line effect a hierarchical clustering analysis on differentially expressed miRNAs in control cells was performed.The statistical analysis underlying the characterization as *differentially expressed* is shown as follows: 


### 1.3.1 Checking assumptions for statistical tests
  

```{r}
# check for outliers 
outl <- dat %>% 
  group_by(cell_line,Irradiation, miRNA) %>%
  identify_outliers(log_exp) 
any(outl$is.extreme)
```

No extreme outliers were present in the analysis. Testing for normality or homogeneity of variances was not indicated in this study as the sample size was too small (n = 4) to detect deviations from said assumptions. Therefore,  a theoretical approach has been used to solve this problem:

* As explained in section 1.1, expression data follows a log-normal distribution, thus log-transformed values were used for statistical analysis 
* To deal with possible heterogeneity of variances, ANOVA with heteroscedasticity corrected covariance matrices (White 1980, Long and Ervin 2000) and post-hoc pairwise t-test with unequal variances were employed (white.adjust = TRUE in ANOVA and pool.sd = TRUE in pairwise t-test)

### 1.3.2 Two-Way ANOVA

A two-way ANOVA was conducted at each level of the factor *miRNA* to determine the effects of cell line and irradiation on miRNA expression. Statistical significance was accepted for p < 0.05 after fdr-adjustment (Benjamini & Hochberg 1995) for two-way interactions and simple main effects. The results of the two-way ANOVA are shown exemplary below.

STR-SHIFT + Enter --> code chunk + output
```{r, message=FALSE}
# 2-way ANOVA with miRNA as moderator variable
two_way_ANOVA <- dat %>%
  group_by(miRNA) %>%
  # perform ANOVA with hccm correction (white.adjust = TRUE)
  anova_test(log_exp ~ cell_line*Irradiation, white.adjust = T) %>%
  adjust_pvalue(method = "fdr") 
two_way_ANOVA
```

In 5 out of the 40 miRNAs there was an interaction effect present (miR-126-3p, miR-146a-5p, miR-30a-3p, miR-30b-5p, miR-7-5p). A cell line main effect was observed in 38 out of 40 miRNAs (each of the miRNAs showing an interaction effect also showed a sigfnifcant main effect). Only for miRNA let-7f-p and miR-30d-5p no cell line effect could be observed after adjusting for multiple testing. 


```{r}
# show interaction
two_way_interaction <- two_way_ANOVA %>%
  filter(Effect == "cell_line:Irradiation" & p.adj < 0.05) %>%
  pull(miRNA)
two_way_ANOVA %>% 
  filter(miRNA %in% two_way_interaction & Effect == "cell_line:Irradiation") 

# show main cell line effect
two_way_cell<- two_way_ANOVA %>%
  filter(Effect == "cell_line" & p.adj < 0.05) %>%
  pull(miRNA)
two_way_ANOVA %>% 
  filter(miRNA %in% two_way_cell & Effect == "cell_line") 

# no effect
two_way_ANOVA %>% 
  filter(!miRNA %in% two_way_cell & Effect == "cell_line")
```

### 1.3.3 One-Way ANOVA

To further investigate the cell line effect a one-way ANOVA was conducted with miRNA and Irradiation as grouping variables. To eliminate the Irradiation effect and only compare *baseline* miRNA expression between cell lines the results were filtered for control cells only, resulting in 36 differentially expressed miRNAs.


```{r, message=FALSE}
# 1-way ANOVA 
one_way_ANOVA <- dat %>%
  group_by(miRNA, Irradiation) %>%
  # perform ANOVA with hccm correction (white.adjust = TRUE)
  anova_test(log_exp ~ cell_line, white.adjust = TRUE) %>%
  adjust_pvalue(method = "fdr") %>% 
  filter(Irradiation == "control" & p.adj < 0.05) %>%
  select(-Irradiation)
one_way_ANOVA
```

*NOTE: using the pooled variance from a 2-way ANOVA in the following 1-way ANOVA yields a higher statistical power but is more susceptible to violations from the homogeneity of variance assumptions. Analogously post-hoc tests with pooled variances (pool.sd = TRUE) are better in situations when the assumptions are met, at the cost of being less robust when assumptions not met.*

### 1.3.4 Cluster analysis

The differentially expressed miRNAs were used in a combined approach using supervised kmeans to cluster cell lines (columns) and unsupervised hierachical clustering to cluster miRNAs (rows). Furthermore hierarchical clustering was also applied in the subclusters split by kmeans. The results were visualized in a Heatmap to reduce complexity and show the quantitative properties of each miRNA as an overview (Fig. 4). 


```{r}
# Heatmap  
# transform data for Heatmap format
dat_Heatmap <- dat %>% 
  filter(Irradiation == "control" & miRNA %in% one_way_ANOVA$miRNA) %>%
  select(-c(Irradiation, log_exp)) %>%
  spread(miRNA, expression) %>%
  mutate(cell_line = str_replace_all(cell_line, "_","-"))

# scale data for Heatmap
dat_scaled <- dat_Heatmap %>% 
  column_to_rownames("ID") %>% 
  select(-cell_line) %>%
  scale() %>%
  t()

# set ID as rownames so colorbar works properly
dat_colorbar <- dat_Heatmap %>% select(c(ID, cell_line)) %>% 
  column_to_rownames("ID")

# define colors for colorbar to identify cell lines
colorbar <- HeatmapAnnotation(
  df =dat_colorbar,
  col = list(
    cell_line = c(
      "Met-1" = "#0073C2FF",
      "Met-4" = "#EFC000FF",
      "SCC-12"="#868686FF" ,
      "SCC-13"="#CD534CFF",
      "SCL-II"="#7AA6DCFF")
    ),
  annotation_legend_param = list(
    cell_line = list(nrow=1)
    )
  )

# define color for the Heatmap cells
col_fun = colorRamp2(c(-1.8, 0, 1.8), c(c("#6D9EC1", "white", "#E46726")))

# Draw Heatmap 
Ht <- Heatmap(
  dat_scaled,
  col= col_fun,
  top_annotation = colorbar,
  row_split = 4,
  column_title = c("A", "B", "C"),
  border = T,
  # split cell lines in 3 clusters according to PCA
  column_km = 3, 
  # repeat km split 100 times to get a stable cluster
  column_km_repeats = 100, 
  # define distance metric and clustering algorithm
  clustering_method_row = "average", # average linkage
  clustering_method_columns = "average",
  clustering_distance_row = "pearson", 
  clustering_distance_column = "euclidean",
  # plot cosmetics 
  rect_gp = gpar(col = "white",lty = 1, lwd = 1),
  row_names_gp = gpar(fontsize = 10),
  show_column_names = FALSE,
  column_names_gp = gpar(fontsize = 10),
  heatmap_legend_param = list(
    title = "row Z-score",
    at = seq(-2,2,by=1),
    color_bar="continuous",
    title_position ="topcenter",
    legend_direction = "horizontal",
    legend_width = unit(4, "cm")
    ))

# plot the final heatmap with annotation and legend
draw(Ht, annotation_legend_side = "bottom", heatmap_legend_side = "bottom")
```

![**Fig. 4: Hierarchical clustering analysis of differentially expressed miRNAs between the five different cell lines.** K-means clustering was applied with k=3 splits as PCA indicated 3 distinct clusters (Met-1/Met-4, SCC-12/SCC-13 and SCL-II). These clusters are also present in the Heatmap displayed by the color code below the column dendrogram. Orange shows upregulation of miRNA expression, blue shows downregulation of miRNA expression. Clustering method: average linkage. Distance measure for miRNAs: pearson correlation. Distance measure for cell lines: euclidean distance. n = 4. )](https://raw.githubusercontent.com/MBender1992/PhD/Marc/Results/Heatmap_cell_lines.png)

The combined cluster analysis showed a clear separation of miRNA expression depending on the cell line context (A: SCC-12/SCC-13,B: SCL-II,C: Met-1/Met-4).
The miRNAs were clustered in 4 groups. Cluster 1 was upregulated in cell line cluster A compared to cluster B and C. Cluster 2 was upregulated in cell line cluster B compared to cluster A and C. Cluster 4 was upregulated in cell line cluster C compared to A and B. Cluster 3 showed an intermediate expression of miRNAs in Cluster B and C with the lowest expression values observed in Cluster A.

Taken together these findings indicate distinct miRNA clusters representing a cell line specific epigenetic pattern and possibly reflecting differentiation, p53 status or other phenotypes associated with these patterns. To investigate the biological function of the miRNAs in each cluster an overrepresentation analysis was conducted to find enriched pathways targeted by these miRNAs. 

**TABELLE MIT PATHWAYS EINFÜGEN**



## 1.4 Investigation of the irradiation effect

To examine the effect of the chronic Irradiation and identify in which cell lines there was an irradiation effect pairwise comparisons between treated and control samples were conducted for each cell line using t-tests. To balance Type I and Type II error rate the results were only adjusted for multiple testing at the ANOVA (section 1.3.2) level using fdr-correction. Subsequent t-test were not adjusted for multiple testing as suggested by Feise (2002) and Rothman (1990). Confirmation of the results was rather achieved by a follow-up qPCR (Fig. 7). 

Differential expression was accepted when fdr(ANOVA) < 0.05, p(t-test) < 0.05 and a fold-change > 1.5 between irradiated cells and controls was observed. Analysis of miRNAs was split into two parts: examination of the interaction effect and the irradiation main effect. 


### 1.4.1 Interaction effect of cell line and irradiation

As mentioned above 5 miRNAs showed a significant interaction effect: 


```{r}
# miRNAs with a significant interaction effect
two_way_ANOVA %>% 
  filter(miRNA %in% two_way_interaction & Effect == "cell_line:Irradiation") 
```

Data was summarized using geometric mean and geometric sd to show data in the original scale but summarized with respect to the log-normal distribution of the miRNA expression values. Fold changes were calculated by dividing the geometric mean of the irradiated samples by the geometric mean of the control cells for each miRNA and cell line, respectively. 
For easier computation and symmetry around 0, fold changes were log2 transformed (log2FC).The analogous threshold to FC > 1.5 is  |log2FC| >  0.585. 

```{r, message = FALSE}
# summarize data with geometric mean and geometric sd for each miRNA in each cell line for irradiated cells and controls
dat_summary <- dat %>% 
  mutate(expression = expression + 1) %>%
  group_by(cell_line, Irradiation,miRNA) %>%
  summarize(geomean = geoMean(expression, na.rm =T), geosd = geoSD(expression))
# calculate fold changes by dividing the geomean expression value of irradiated cells by geomean expression value of 
# control cells for each each miRNA in each cell line
folds <-  dat_summary %>% 
  group_by(cell_line,  miRNA) %>% 
  summarize(FC = geomean[Irradiation == "KAUVIR"]/geomean[Irradiation == "control"]) %>% 
  mutate(log2FC = log2(FC)) %>%
  setNames(c("cell_line", "miRNA","FC", "log2FC"))
```

Combining the criterea for different expression as described in section 1.4, four miRNAs remained (Fig. 5).

```{r, message = FALSE}
# t-test for the interaction effect
pwc_interaction <- dat %>%
  group_by(miRNA,cell_line) %>%
# set pool.sd = F to use different variances for calculations
  pairwise_t_test(log_exp~Irradiation, pool.sd = F) %>%
#filter only miRNAs with p < 0.05 and a significant interaction term in two way ANOVA
  filter(p < 0.05 & miRNA %in% two_way_interaction)  


#calculate differential expression for interaction
diff_expr_interaction <- pwc_interaction %>%
  inner_join(folds) %>%
  select(miRNA, cell_line, p, log2FC) %>% 
  filter(abs(log2FC) > log2(1.5))
diff_expr_interaction
```

```{r}
# function expands facet limits individually to plot each facet with an individual scale
facet_limits <-function(data, y, group){
  
  dat <- data.table(data)
  
  dat[,y_min := eval(parse(text = y))*0.5, by = eval(parse(text = group))]
  dat[,y_max:= eval(parse(text = y))*1.25, by = eval(parse(text = group))]
  
  dat <- as_tibble(dat)
  return(dat)
}

# apply function facet_limits to generate individual facet limits for the plot
dat_plot <- facet_limits(dat_summary, "geomean + geosd", "miRNA") %>% 
  mutate(cell_line = str_replace_all(cell_line, "_","-"))

# define limits of errorbar to only show upper errorbar
limits <- aes(ymax = ifelse(geomean>0,geomean + geosd,geomean/2),
              ymin = ifelse(geomean<0,geomean - geosd,geomean/2))


# generating Plot
dat_plot %>% 
  # only show differentially expressed miRNAs with a significant interaction term in two-way ANOVA
  filter(miRNA %in% diff_expr_interaction$miRNA) %>%
  ggplot(
    aes(
      x=cell_line,
      y=geomean,
      fill=Irradiation
      )
    ) + 
  # add errorbars with the defined limits
  geom_errorbar(
    limits,
    width=.2,
    position=position_dodge(0.5)
    ) +
  # add barplots to show geometric mean
  geom_bar(
    stat="identity",
    color="black",
    position=position_dodge(),
    width= 0.5
    ) +
  # split plot into separate plot for each miRNA and change theme
  facet_wrap(~miRNA,scales="free")+ 
  theme_PhD(axis.text.size = 8) +
  theme(
    axis.title.x = element_blank(),
    strip.background = element_blank(),
    legend.key.size = (unit(0.7,"line")),
    legend.position = "bottom"
  ) +
  # extend facets by adding blank datapoints as defined with the function facet limits
  geom_blank(aes(y = y_min)) +
  geom_blank(aes(y = y_max)) + 
  # change colors of the plot elements and add y axis title
  scale_fill_manual(values = grey.colors(5,start=1,end=0.2))  +
  scale_y_continuous(expand = c(0, 0)) +
  ylab("miRNA expression (a.u.)")
```


![**Fig. 5: miRNA expression of the differentially expressed miRNAs with an interaction effect of cell line and irradiation in the two-way ANOVA (section 1.3.2).** KAUVIR: cells irradiated with the irradiation device for the KAUVIR project. Results of interaction term in two-way ANOVA after fdr-adjustment are shown above the plots. Student's t-test was used for group comparisons. &#42;&#42;&#42;: p < 0.001, &#42;&#42;: p < 0.01, &#42;: p < 0.05. geomean ± geosd.  n = 4. ](https://raw.githubusercontent.com/MBender1992/PhD/Marc/Results/miRNA_expression_interaction.png)

All 4 miRNAs were upregulated in at least one cell line (miR-126-3p in SCC-13, miR-146a-5p in SCL-II, miR-30a-3p in SCC-12 and SCL-II and miR-7-5p in SCL-II). Although not significantly changed the same trend of upregulation can be observed for other cell lines as well, with an exception of miR-126-3p in Met-1 and SCC-12, miR-146a-5p in SCC-12 and SCC-13 and miR-7-5p in Met-1 and SCC-13 (which even shows a downregulation after irradiation, however not significantly). 



### 1.4.2 Main effect Irradiation

Subsequently the irradiation main effect has been investigated to analyze miRNAs with a common UV-response in the different cell types (Fig. 6). 

```{r, message = FALSE}
# filter miRNAs with a significant irradiation term but without significant interaction term
two_way_irradiation <- two_way_ANOVA %>%
  filter(Effect == "Irradiation" & p.adj < 0.05 & !miRNA %in% two_way_interaction) %>%
  pull(miRNA)

# t-test for the irradiation effect
pwc_irradiation <- dat%>%
  group_by(miRNA,cell_line) %>%
  pairwise_t_test(log_exp~Irradiation, pool.sd = F) %>% 
  filter(p < 0.05 & miRNA %in% two_way_irradiation) 

#calculate differential expression for Irradiation
diff_expr_irradiation <- pwc_irradiation %>%
  inner_join(folds) %>%
  select(miRNA, cell_line, p, log2FC) %>% 
  filter(abs(log2FC) > log2(1.5))
diff_expr_irradiation
```


```{r}
dat_plot %>% 
  # only show differentially expressed miRNAs with a significant main effect "irradiation" term in two-way ANOVA but without significant interaction
  filter(miRNA %in% diff_expr_Irradiation$miRNA) %>%
  ggplot(
    aes(
      x=cell_line,
      y=geomean,
      fill=Irradiation
    )
  ) + 
  # add errorbars with the defined limits
  geom_errorbar(
    limits,
    width=.2,
    position=position_dodge(0.5)
  ) +
  # add barplots to show geometric mean
  geom_bar(
    stat="identity",
    color="black",
    position=position_dodge(),
    width= 0.5
  ) +
  # split plot into separate plot for each miRNA and change theme
  facet_wrap(~miRNA,scales="free")+ 
  theme_PhD(axis.text.size = 8) +
  theme(
    axis.title.x = element_blank(),
    strip.background = element_blank(),
    legend.key.size = (unit(0.7,"line")),
    legend.position = "bottom"
  ) +
  # extend facets by adding blank datapoints as defined with the function facet limits
  geom_blank(aes(y = y_min)) +
  geom_blank(aes(y = y_max)) + 
  # change colors of the plot elements and add y axis title
  scale_fill_manual(values = grey.colors(5,start=1,end=0.2))  +
  scale_y_continuous(expand = c(0, 0)) +
  ylab("miRNA expression (a.u.)")
```

![**Fig. 6: miRNA expression of the differentially expressed miRNAs with an irradiation main effect in the two-way ANOVA but without significant interaction term (section 1.3.2).** KAUVIR: cells irradiated with the irradiation device for the KAUVIR project. Results of irradiation term in two-way ANOVA after fdr-adjustment are shown above the plots. Student's t-test was used for group comparisons. &#42;&#42;: p < 0.01, &#42;: p < 0.05. geomean ± geosd.  n = 4. ](https://raw.githubusercontent.com/MBender1992/PhD/Marc/Results/miRNA_expression_irradiation.png)

Significant upregulation could be detected for miR-135b-5p in SCC-13, miR-183-5p in SCC-13, miR-200a-3p in SCC-12, miR-30d-5p in SCC-13 and miR-424-5p in SCC-13. The miRNA expression shows a similar trend in the other cell lines as well with an exception of miR-135b-5p in Met-1 and SCL-II, miR-200a-3p in SCC-13 and miR-424-5p in Met-1. miR-205-5p is the only miRNA exhibiting a downregulation in all cell lines, with a significant downregulation in SCL-II. Furthermore, miR-205-5p shows the most significant general irradiation effect with p = 0.000253 in the two-way ANOVA. In summary UV irradiation seems to induce miRNA expression more often than inhibting miRNA expression with miR-205-5p being the only significantly downregulated miRNA out of totally 10 differentially expressed miRNAs between treated and control samples (results from section 1.4.1 and 1.4.2 combined). 

Intriguingly, significant irradiation effects were only present in the three cell lines SCC-12, SCC-13 and SCL-II. This effect has been further investigated in section 1.5. 

### 1.4.3 qPCR Validation

To validate the findings of section 1.4.1 and section 1.4.2 qPCR was performed.  


##  1.5 UV sensitivity of cell lines